---
title: "Lab 3 Assignment - Scale"
output: html_notebook
---
## Challenge 1 (4 points)

**Build a raster with 100 rows and 100 columns. Fill the raster cells with values of a random variable drawn from a distribution of your choosing (Poisson, Normal, Uniform, etc.). Calculate the mean and variance of the values in that raster. Now increase the grain size of those cells by factors of 2, 5, and 10, combining cell values using a mean function. At each iteration, calculate the mean and variance of the values in the resulting raster. Generate 2 scatterplots that have grain size on the x-axis. Plot the mean raster value on the y-axis of the first, and variance on the y-axis of the second. What do you notice about how these values change as you "scale up" the grain size? Why do you think this pattern occurs?**

Place your code in the code chunk below so I can reproduce your analyses/figures.

```{r}
challenge1<-rast(ncol=100, nrow=100, xmin=1, xmax=5, ymin=1, ymax=5)
challenge1[] = rpois(ncell(challenge1), lambda=3)
plot(challenge1)
mean(as.matrix(challenge1))
var(as.matrix(challenge1))

challenge1Fact2<-aggregate(challenge1, fact=2, fun='mean')
plot(challenge1Fact2)
mean(as.matrix(challenge1Fact2))
var(as.matrix(challenge1Fact2))

challenge1Fact5<-aggregate(challenge1, fact=5, fun='mean')
plot(challenge1Fact5)
mean(as.matrix(challenge1Fact5))
var(as.matrix(challenge1Fact5))

challenge1Fact10<-aggregate(challenge1, fact=10, fun='mean')
plot(challenge1Fact10)
mean(as.matrix(challenge1Fact10))
var(as.matrix(challenge1Fact10))

grain<-c(1, 2, 5, 10)
mean_values<-c(mean(as.matrix(challenge1)), mean(as.matrix(challenge1Fact2)), mean(as.matrix(challenge1Fact5)), mean(as.matrix(challenge1Fact10)))
plot(mean_values~grain)

variance_values<-c(var(as.matrix(challenge1)), var(as.matrix(challenge1Fact2)), var(as.matrix(challenge1Fact5)), var(as.matrix(challenge1Fact10)))
plot(variance_values~grain)
```

As the scale increases, the mean stays the same but the variance decreases. This is because the heterogeneity is effectively averaged out, or masked by the average. 


## Challenge 2 (4 points)

**Identify a situation in which you might use a summary function other than the mean to calculate new cell values when you scale up the grain of a raster (e.g., median, mode, minimum, maximum, etc.). Repeat the effort from Challenge 1 using this alternate function. Again, create two scatterplots showing how the mean and variance values of the raster change as you scale up the cell size by factors of 2, 5, and 10. Do you see a similar pattern? Compare and contrast your findings with those from Challenge 1.**

*Hint: You should be able to recycle your code from Challenge 1 with only a couple of small tweaks to answer this question.*

Place your code in the code chunk below so I can reproduce your analyses/figures.

```{r}
challenge2<-rast(ncol=100, nrow=100, xmin=1, xmax=5, ymin=1, ymax=5)
challenge2[] = rpois(ncell(challenge1), lambda=3)
plot(challenge2)
summary(challenge2)

challenge2Fact2<-aggregate(challenge1, fact=2, fun='mean')
plot(challenge2Fact2)
summary(challenge2Fact2)

challenge2Fact5<-aggregate(challenge1, fact=5, fun='mean')
plot(challenge2Fact5)
summary(challenge2Fact5)

challenge2Fact10<-aggregate(challenge1, fact=10, fun='mean')
plot(challenge2Fact10)
summary(challenge2Fact10)

grain<-c(1, 2, 5, 10)
meanChallenge2<-c(mean(as.matrix(challenge2)), mean(as.matrix(challenge2Fact2)), 
                 mean(as.matrix(challenge2Fact5)), mean(as.matrix(challenge2Fact10)))
plot(meanChallenge2~grain)

varianceChallenge2<-c(var(as.matrix(challenge2)), var(as.matrix(challenge2Fact2)), 
                     var(as.matrix(challenge2Fact5)), var(as.matrix(challenge2Fact10)))
plot(varianceChallenge2~grain)
```

Again, the mean remains mostly the same, although you can see a slight increase that wasn't apparent in the previous method. You can also see that the range of values decreases - i.e. the maximum decreases and the minimum increases, which wasn't visible before. 

## Challenge 3 (2 points)

**Recall that before we calculated forest cover, we cropped our NLCD raster to minimize its size and the computing effort necessary from our poor little computers. How might that affect our ability to evaluate the scale at which five-lined skinks respond to forest cover? Why?**

Cropping the NLCD raster reduces the extent of the study area and hypothetically should allow us to refine the grain at which we're evaluating skink responses. We might get a better picture of spatial variance when looking at a smaller area because heterogeneity is not averaged out in the larger grain that is typically associated with greater extents.

## Challenge 4 (4 points)

**In the lab, we measured forest cover at 1 km and 5 km. Extract forest cover proportions around each sample point for 100 m, 500 m, 1 km, 2 km, 3 km, 4 km, and 5 km scales. Examine the correlation between these 7 variables (remember the chart.Correlation() function). What patterns do you notice in correlation among these variables?**

*Hint: Recall the for loop we used to calculate this variable at two scales... could you make a small addition here to look at more scales?*

```{r}
sites = st_read("/vsicurl/https://github.com/ValenteJJ/SpatialEcology/raw/main/Week3/reptiledata.shp") %>% 
  filter(management!='Corn')
st_crs(sites) = "+proj=aea +lat_0=23 +lon_0=-96 +lat_1=29.5 +lat_2=45.5 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
head(sites)

presAbs = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week3/reptiles_flsk.csv')

sites = sites %>% 
  left_join(presAbs, by='site')

studyArea = st_bbox(sites) + c(-10000, -10000, 10000, 10000)
studyArea = st_as_sfc(studyArea)

nlcd = get_nlcd(studyArea,
                label='studyArea',
                year = 2016,
                dataset = 'landcover',
                landmass = 'L48'
)

forest = nlcd %>% 
  setValues(0)

forest[nlcd=='Deciduous Forest' | nlcd=='Evergreen Forest' | nlcd=='Mixed Forest'] = 1

bufferCover<-function(shp, size, landcover){
  buffArea<-(pi*size^2)/10000
  grainArea<-(prod(res(landcover)))/10000
  
  buffi<-st_buffer(shp[i,], dist=size)
  cropi<-crop(landcover, buffi, mask=T)
  numCells<-global(cropi, 'sum', na.rm=T)
  forestHa<-numCells * grainArea
  propForest<-forestHa / buffArea
  
  return(propForest)
}

for100m<-as.vector(rep(NA, nrow(sites)))
for500m<-as.vector(rep(NA, nrow(sites)))
for1km<-as.vector(rep(NA, nrow(sites)))
for2km<-as.vector(rep(NA, nrow(sites)))
for3km<-as.vector(rep(NA, nrow(sites)))
for4km<-as.vector(rep(NA, nrow(sites)))
for5km<-as.vector(rep(NA, nrow(sites)))

for(i in 1:nrow(sites)){
  for100m[i] = bufferCover(sites, 100, forest)
  for500m[i] = bufferCover(sites, 500, forest)
  for1km[i] = bufferCover(sites, 1000, forest)
  for2km[i] = bufferCover(sites, 2000, forest)
  for3km[i] = bufferCover(sites, 3000, forest)
  for4km[i] = bufferCover(sites, 4000, forest)
  for5km[i] = bufferCover(sites, 5000, forest)
}

forestData = sites %>% 
  mutate(for100m = unlist(for100m),
         for500m = unlist(for500m),
         for1km = unlist(for1km),
         for2km = unlist(for2km),
         for3km = unlist(for3km),
         for4km = unlist(for4km),
         for5km = unlist(for5km))

head(forestData)

forestData %>% 
  as.data.frame() %>% 
  select(coords_x1, for100m, for500m, for1km, for2km, for3km, for4km, for5km) %>% 
  PerformanceAnalytics::chart.Correlation(histogram=F)
```
I'm not sure that I'm interpreting this correctly, but what I'm gathering is a positive correlation that increases in strength as scale increases. 

## Challenge 5 (4 points)

**Fit 8 logistic regression models (a null model and one for each of the 7 forest scales). Compare these models using AICc. Which scale do you think represents the critical or characteristic scale at which forest cover affects skink presence? Is this scale clearly better than the others, or is there some ambiguity? What are some mechanisms by which forest cover could affect skink presence at this scale? What is your overall conclusion regarding how forest cover affects skink presence (i.e., take a look at the betas)?**

Place your R code in the chunk below.
```{r}
modelNull<-glm(pres~1, family='binomial', data=forestData)
model100m<-glm(pres~for100m, family='binomial', data=forestData)
model500m<-glm(pres~for500m, family='binomial', data=forestData)
model1km<-glm(pres~for1km, family='binomial', data=forestData)
model2km<-glm(pres~for2km, family='binomial', data=forestData)
model3km<-glm(pres~for3km, family='binomial', data=forestData)
model4km<-glm(pres~for4km, family='binomial', data=forestData)
model5km<-glm(pres~for5km, family='binomial', data=forestData)

aictab(list(modelNull, model100m, model500m, model1km, model2km, model3km, model4km, model5km), modnames=c('Null', '100 m', '500 m', '1 km', '2 km', '3 km', '4 km', '5 km'))
```

The AIC scores suggest that the 2 km model is the best fit among the models considered. However, the delta AIC for the second-ranked 4 km model is low (0.36). And the third- and fourth-ranked models (3 km and 5 km, respectively) are still less than 2. (Related: I'm not sure if that matters here because all of the models have the same number of parameters.) Regardless of the value of the 3 km and 5 km models, the 2 km and 4 km models are negligibly different, making it difficult to determine which of the two is actually best. 

I think the relative availability of edge habitat might affect skinks at this scale. It's a 2D version of the surface area:size thing. If there is some quality of the edge that skinks need or benefit from (maybe it's easier to find sunshine to bask in), smaller patches of forest might have *too* much edge for their size. Perhaps there's not enough water/food/cover inside such a small area. Large patches, on the other hand, may force the animal to travel relatively long distances to access internal and edge resources and features. 

My overall conclusion is that it may be a range of scales. Although, the validity of this probably depends on how much stock you put in the 3 km model. 

## Challenge 6 (2 points)

**If you encounter ambiguity in identifying the characteristic scale of an effect, can you come up with a clever way to condense the information in the multi-scale variables into just one or two? When might it be ok to include two covariates in the same model (think multiple regression) that represent the same ecological feature measured at different scales (e.g., forest cover at 1 km AND forest cover at 5 km in the same model)? I can think of both a biological and a statistical answer to this question.**

I think you could include two covariates for the same feature if one affected organisms at the individual scale and one affected organisms at a population scale. There could be processes at the 1 km scale that affect whether an individual skink might momentarily choose one patch of habitat over another, as well as processes at the 5 km scale that determine if a patch of habitat can sustain skinks over time. 

In terms of resolving this statistically, maybe you could nest the effects of smaller scales and larger scales. I don't know if this is clever or if it would work but it's all I can think of.
\
\
\
\


