---
title: "Species Distribution Modeling"
output: html_notebook
---

```{r}
require(tidyterra)
require(dismo)
require(tidyverse)
require(terra)
require(predicts)
require(ggnewscale)
require(mgcv)
require(randomForest)
require(maxnet)
require(enmSdmX)
require(gbm)
require(landscapemetrics)
```

```{r}
# read in the count data
vathData <- read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week8/vath_2004.csv')

# create subsets
vathPres <- vathData %>%
  filter(VATH==1)
vathAbs <- vathData %>%
  filter(VATH==0)

# coords only - no count
vathPresXY <- as.matrix(vathPres %>%
                          select(EASTING, NORTHING))
vathAbsXY <- as.matrix(vathAbs %>%
                         select(EASTING, NORTHING))

# read in the covariate data
elev <- rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week8/elevation.tif')
canopy <- rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week8/canopy.tif')
mesic <- rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week8/mesic.tif')
precip <- rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week8/precip.tif')

# project the rasters
crs(elev) = crs(mesic)
crs(canopy) = crs (mesic)

# resample the rasters
mesic <- resample(x=mesic, y=elev, 'near') 
precip <- resample(x=precip, y=elev, 'bilinear')

# extract mesic forest values within 1 km of sampling locations (note to self: scale shows strong correlations of avian distribution, choice based on literature)
probMatrix <- focalMat(mesic, 1000, type='circle', fillNA=FALSE)
mesic1km <- focal(mesic, probMatrix, fun='sum')

# combine
layers <- c(canopy, elev, mesic1km, precip)
names(layers) = c('canopy', 'elev', 'mesic1km', 'precip')

# creating the background points
set.seed(23)
backXY <- data.frame(backgroundSample(layers, n=2000, p=vathPresXY))

# create dataframes containing all of the information for each point
presCovs <- terra::extract(layers, vathPresXY)
backCovs <- terra::extract(layers, backXY)

presCovs <- data.frame(vathPresXY, presCovs, pres=1)
backCovs <- data.frame(backXY, backCovs, pres=0)

# and remove sites without corresponding environmental data
presCovs <- presCovs[complete.cases(presCovs),]
backCovs <- backCovs[complete.cases(backCovs),]

# combine the remaining data
backCovs <- backCovs %>%
  select(-ID)
colnames(presCovs)[1:2]=c('x','y')

presBackCovs <- rbind(presCovs, backCovs)
```



# Challenge 1 (4 points)

In the lab, we created 6 species distribution models (SDMs) for the same species using 6 different techniques. Plot the maps generated from (1) the bioclim envelope function, (2) the GLM model, and (3) the random forest model next to one another. What similarities and differences do you notice among these maps? What might explain some of these differences?

```{r}
# bioclim envelope
temp <- presCovs %>%
  select(elev, precip, mesic1km, canopy) %>%
  as.matrix()

bioclimSDM <- envelope(temp)
bioclimMap <- predict(layers, bioclimSDM)

# GLM
glmBack <- glm(pres~canopy+elev+I(elev^2)+mesic1km+precip, data=presBackCovs, family = 'binomial') # note to self: I(elev^2) allows for a non-linear/quadratic relationship because thrushes may be most likely to occur at moderate elevations

glmMap <- predict(layers, glmBack, type='response') # type='response' to make predictions on the probability scale (otherwise on the link scale)

# random forest
# tuneRF(y=as.factor(presBackCovs$pres), x=presBackCovs[,3:6], stepFactor=2, nTreeTry=500)
rfModel <- randomForest(as.factor(pres)~canopy+elev+mesic1km+precip, data=presBackCovs, mtry=2, ntree=500, na.action=na.omit)
rfMap <- predict(layers, rfModel, type='prob', index=2) # index=0 because there are two classes to make predictions from (0 and 1; i.e. absence and presence)

# plots combined (why ugly?)
par(mfrow = c(1, 3))
plot(bioclimMap)
title('bioclim SDM')
plot(glmMap)
title('glm SDM')
plot(rfMap)
title('random forest SDM')
```
The maps are similar for the most part in terms of the general areas that are identified as potentially being inhabited by the birds.

I think the envelope model predicts the greatest overall quantity of high values. This makes sense, since envelope models demonstrate the degree to which locations fall within the variation of observed locations, which tends to result in over-predictions.

The glm SDM is interesting in that it accounts for the quadratic relationship between elevation and thrushes, which I don't think that the other two models do (?). I do think that it increases the amount of land that satisfies the elevation criteria and, by extension, the amount of pale pink/lower value area.

At a glance, the random forest SDM predicts the least amount of non-zero values overall.

# Challenge 2 (4 points)

When we fit our GLM in lab, we used background points, rather than true absence points, to represent pseudo-absences. Fit the exact same GLM model, only this time use presence and true absence data. That is, replace the background rows in the dataframe with rows that represent actual sites where surveys were completed but Varied Thrush were not detected. Once you've fit the GLM, build a new SDM from this fitted model and visually compare the prediction surface to that built based on the presence-background model. What discrepancies do you notice, and what is your intuition regarding which of these two models is more reliable?

```{r}
# replace the background rows in the dataframe with rows that represent true absences from the initial survey

trueAbsCovs <- terra::extract(layers, vathAbsXY) # extract the values for each location where thrushes were not observed
trueAbsCovs <- data.frame(vathAbsXY, trueAbsCovs, pres=0) # merge them with absence classification (i.e. 0)
trueAbsCovs <- trueAbsCovs[complete.cases(trueAbsCovs),] # remove sites where environmental data is nonexistent/incomplete
colnames(trueAbsCovs)[1:2]=c('x','y') # adjust to match presCovs
presTrueAbsCovs <- rbind(presCovs, trueAbsCovs) # merge with the presence data

# glm with presence and true absence data
glmTrue <- glm(pres~canopy+elev+I(elev^2)+mesic1km+precip, data=presTrueAbsCovs, family = 'binomial')

# new SDM 
glmTrueMap <- predict(layers, glmTrue, type='response')

# compare the models
par(mfrow = c(1, 2))
plot(glmMap)
title('background points')
plot(glmTrueMap)
title('true absences')
```
The model based on true absences depicts values all the way up to 1, while the model based on background points is still depressed, which I now know to be an artifact of the number of absence points relative to presence points. I think it comes down to whether or not potentially imperfect absence data is more credible than simulated absence data in terms of the model's ability to make accurate predictions and I really don't know the answer to that.

# Challenge 3 (4 points)

Now plot the relationship between the 4 explanatory variables and the predicted occupancy values based on the two fitted GLM models (presence-background and presence-absence). Recall that we did this in the latter part of our lab. Do you notice any differences in the covariate patterns between the two models? Does this help you interpret the discrepancies between the predicted surfaces from the two models?

```{r}
# create medians
canopyMedian <- median(backCovs$canopy)
precipMedian <- median(backCovs$precip)
mesic1kmMedian <- median(backCovs$mesic1km)
elevMedian <- median(backCovs$elev)

# elevation
elevRange <- seq(min(backCovs$elev), max(backCovs$elev), length=1000)
elevTmp <- data.frame(expand.grid(elev=elevRange, canopy=canopyMedian, precip=precipMedian, mesic1km=mesic1kmMedian))
elevData <- data.frame(glmBackElev = predict(glmBack, elevTmp, type='response'),
                       glmTrueElev = predict(glmTrue, elevTmp, type='response')) %>%
  cbind(elevTmp) %>%
  select(glmBackElev:elev) %>%
  pivot_longer(cols = c(glmBackElev:glmTrueElev), names_to = "model") %>%
  mutate(variable='elevation')

# precipitation
precipRange <- seq(min(backCovs$precip), max(backCovs$precip), length=1000)
precipTmp <- data.frame(expand.grid(precip=precipRange, canopy=canopyMedian, elev=elevMedian, mesic1km=mesic1kmMedian))
precipData <- data.frame(glmBackPrecip = predict(glmBack, precipTmp, type='response'),
                       glmTruePrecip = predict(glmTrue, precipTmp, type='response')) %>%
  cbind(precipTmp) %>%
  select(glmBackPrecip:precip) %>%
  pivot_longer(cols = c(glmBackPrecip:glmTruePrecip), names_to = "model") %>%
  mutate(variable='precipitation')

# canopy
canopyRange <- seq(min(backCovs$canopy), max(backCovs$canopy), length=1000)
canopyTmp <- data.frame(expand.grid(canopy=canopyRange, elev=elevMedian, precip=precipMedian, mesic1km=mesic1kmMedian))
canopyData <- data.frame(glmBackCanopy = predict(glmBack, canopyTmp, type='response'),
                       glmTrueCanopy = predict(glmTrue, canopyTmp, type='response')) %>%
  cbind(canopyTmp) %>%
  select(glmBackCanopy:canopy) %>%
  pivot_longer(cols = c(glmBackCanopy:glmTrueCanopy), names_to = "model") %>%
  mutate(variable='canopy')

# mesic forest
mesicRange <- seq(min(backCovs$mesic1km), max(backCovs$mesic1km), length=1000)
mesicTmp <- data.frame(expand.grid(mesic1km=mesicRange, canopy=canopyMedian, precip=precipMedian, elev=elevMedian))
mesicData <- data.frame(glmBackMesic = predict(glmBack, mesicTmp, type='response'),
                        glmTrueMesic = predict(glmTrue, mesicTmp, type='response')) %>%
  cbind(mesicTmp) %>%
  select(glmBackMesic:mesic1km) %>%
  pivot_longer(cols = c(glmBackMesic, glmTrueMesic), names_to = "model") %>%
  mutate(variable = 'mesic')

# partial plots
colnames(elevData)[1] = colnames(canopyData)[1] = colnames(precipData)[1] = colnames(mesicData)[1] = 'xValue'

tmpPlots <- rbind(elevData, canopyData, precipData, mesicData)

ggplot(tmpPlots, aes(x=xValue, y=value, color=model))+
  facet_wrap(~variable, scales='free_x')+
  geom_line()+
  theme_bw()+
  theme(panel.grid=element_blank())

```
The model based on the true absence data predicted higher values for all four of the environmental variables than the model based on the background points. In most cases (canopy, elevation, and mesic forest), the difference was subtle. However, the true absence model predicted vastly greater values related to precipitation than the background point model did. I think the overall tendency of the true absence-based model to inflate the values is consistent with the pattern that you can see on the surfaces the models generate. I think it also pertains to the number of background points (2000) vs the number of true absences (710). It's hard to draw comparisons between the values because they're influenced by the number of presence points relative to the number of absence points, which obviously differs between the models.

# Challenge 4 (4 points)

Varied Thrush are considered forest-dependent, and thus one might characterize mesic forests as "habitat" for the species. Calculate the total amount of mesic forest in the study area, and the mean size of the mesic forest patches.

Using the SDM built from the random forest model, convert the landscape into "habitat" and "non-habitat." To do this, choose a threshold value in your SDM and convert all cells with predicted outcomes greater than this threshold to 1 and all cells with predicted values below your threshold to 0. Justify your choice of your threshold value. Now calculate the total amount of habitat and mean size of habitat patches based on this new raster (i.e., create patches of "habitat" based on aggregations of cells you deemed 1). How do the habitat amount and patch size values compare between the mesic forest approach and the SDM-based approach? In what situations might you rely on one map over the other?

```{r}
# total amount of mesic forest
mesicTotalArea <- lsm_c_ca(mesic) %>%
  filter(class==1)

# mean size of mesic forest patches 
mesicPatchArea <- lsm_c_area_mn(mesic, directions = 8) %>%
  filter(class==1)

# using the random forest model:
# convert the landscape into "habitat" and "non-habitat"
habitat <- rfMap %>%
setValues(NA)
habitat[rfMap>=0.25]=1 # think of a justification or change the .25 selection? 

# calculate the total amount of habitat
habitatTotalArea <- lsm_c_ca(habitat) %>%
  filter(class==1)

# calculate the mean size of habitat patches
habitatPatchArea <- lsm_c_area_mn(habitat, directions=8) %>%
  filter(class==1)
```

There's substantially more habitat available when you only look at mesic forest metrics. The SDM approach suggests that there's much less habitat truly available when you consider several environmental covariates. I think the landcover-only approach might be sufficient if you're looking at a pretty homogenous landscape. If there's very little variation in elevation or precipitation, for example, then looking at landcover alone may be sufficient. 

# Challenge 5 (4 points)

When we fit the Maxent model in the lab, we used a regularization constant of 1. Fit the model two more times, using regularization (regmult) constants of 0.5 and 3. Construct figures showing the relationship between the 4 explanatory variables and the predicted outcome from these 3 fitted Maxent models. What is the regularization constant doing? Hint: you may need to Google it.

```{r}
# original model
pbVect <- presBackCovs$pres
covs <- presBackCovs %>%
  select(canopy:precip)
maxentModel <- maxnet(p = pbVect,
                      data=covs,
                      regmult=1,
                      classes='lqpht')
# regularization constant of 0.5
maxentModel.5 <- maxnet(p = pbVect,
                      data=covs,
                      regmult=0.5,
                      classes='lqpht')
# regularization constant of 3
maxentModel3 <- maxnet(p = pbVect,
                      data=covs,
                      regmult=3,
                      classes='lqpht')

# partial plots

# elevation
elevData <- data.frame(maxent = predict(maxentModel, elevTmp, type='logistic')[,1],
                       maxent.5 = predict(maxentModel.5, elevTmp, type='logistic')[,1],
                       maxent3 = predict(maxentModel3, elevTmp, type='logistic')[,1]) %>%
  cbind(elevTmp) %>%
  select(maxent:maxent3, elev) %>%
  pivot_longer(cols = c(maxent:maxent3), names_to = "model") %>%
  mutate(variable='elevation')

# precipitation
precipData <- data.frame(maxent = predict(maxentModel, precipTmp, type='logistic')[,1],
                       maxent.5 = predict(maxentModel.5, precipTmp, type='logistic')[,1],
                       maxent3 = predict(maxentModel3, precipTmp, type='logistic')[,1]) %>%
  cbind(precipTmp) %>%
  select(maxent:maxent3, precip) %>%
  pivot_longer(cols = c(maxent:maxent3), names_to = "model") %>%
  mutate(variable='precip')

# canopy
canopyData <- data.frame(maxent = predict(maxentModel, canopyTmp, type='logistic')[,1],
                       maxent.5 = predict(maxentModel.5, canopyTmp, type='logistic')[,1],
                       maxent3 = predict(maxentModel3, canopyTmp, type='logistic')[,1]) %>%
  cbind(canopyTmp) %>%
  select(maxent:maxent3, canopy) %>%
  pivot_longer(cols = c(maxent:maxent3), names_to = "model") %>%
  mutate(variable='canopy')

# mesic forest
mesicData <- data.frame(maxent = predict(maxentModel, mesicTmp, type='logistic')[,1],
                       maxent.5 = predict(maxentModel.5, mesicTmp, type='logistic')[,1],
                       maxent3 = predict(maxentModel3, mesicTmp, type='logistic')[,1]) %>%
  cbind(canopyTmp) %>%
  select(maxent:maxent3, mesic1km) %>%
  pivot_longer(cols = c(maxent:maxent3), names_to = "model") %>%
  mutate(variable='mesic')

# partial plots
colnames(elevData)[1] = colnames(canopyData)[1] = colnames(precipData)[1] = colnames(mesicData)[1] = 'xValue'
tmpPlots <- rbind(elevData, canopyData, precipData, mesicData)
ggplot(tmpPlots, aes(x=xValue, y=value, color=model))+
  facet_wrap(~variable, scales='free_x')+
  geom_line()+
  theme_bw()+
  theme(panel.grid=element_blank())
```
The regularization constant helps balance model complexity and goodness of fit. The lower the regularization constant, the more closely the model is permitted to fit the training data, potentially resulting in overfitting. I think this would be why the maxent.5 lines are oddly shaped. The higher regularization constants smooth out most of the peaks and valleys in the trendline that are artifacts of the actual data. 